Clustering: A pipeline to analyse time-course gene expression data

### 2019-03-17 scripts/00_introduction.Rmd

- TODO why is time-course gene expression analysis important for biology. Why
  are novel workflows necessary.

Time-course gene expression analysis workflows include steps used more broadly
for gene expression data analasys such as quality control (QC), exploratory
data analysis (EDA), pre-processing, normalization, and visualization. Yet, a
typical gene expression data analysis workflow is not entirely adequate for
analysing time-course data. In particular, differential expression analysis
needs to be adapted to time-course data, and often needs to be followed by a
clustering step to group genes sharing similar patterns of expression.

Here, we propose an integrated workflow for downstream analysis of time-course
data, with the following main steps: (1) quality control and normalization;
(2) differential expression analysis of time-course data; (3) clustering; (4)
downstream analysis of clusters with GO term and pathway enrichment.

XXX Add the packages used in the workflow

@@

>>>>

Time-course gene expression analysis workflows include steps used more broadly
for gene expression data analysis such as quality control (QC), exploratory
data analysis (EDA), pre-processing, normalization, and visualization. Yet, a
typical gene expression data analysis workflow is not entirely adequate for
analysing time-course data. In particular, differential expression analysis
needs to be adapted to time-course data, and often needs to be followed by a
clustering step to group genes sharing similar patterns of expression.

Here, we propose an integrated workflow for downstream analysis of time-course
data, with the following main steps: (1) quality control and normalization;
(2) differential expression analysis of time-course data; (3) clustering; (4)
downstream analysis of clusters with GO term, pathway, and motif enrichment.

This workflow relies on several standard packages for analysing gene
expression data, some specific for time-course data, others broadly used by
the community. We also provide alongside this workflow a package that nicely
wraps all calls to dependencies with a common API by defining a model object
holding all necessary metadata to perform all steps of the pipeline.

@@

The authors find that

### 2019-03-17 scripts/03_clustering.Rmd
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

pval_col_to_keep = colnames(de_analysis)[
    grepl("-pval", colnames(de_analysis))]
pvalues = de_analysis[, pval_col_to_keep]
```

@@

The next step is to perform clustering.

@@

kmeans_clusters = moanin:::splines_kmeans(
    y, meta, n_clusters=20,
    random_seed=42,
    n_init=20,
    degrees_of_freedom=df)

# Rescale the centroids
centroids = moanin:::rescale_values(kmeans_clusters$centroids, meta)

@@

```{r fig.height=8, fig.width=8}
moanin:::plot_centroids(centroids, meta,
			colors=as.vector(unlist(ann_colors["Group"])))
```

>>>>

pval_col_to_keep = colnames(de_analysis)[
    grepl("-pval", colnames(de_analysis))]
pvalues = de_analysis[, pval_col_to_keep]

splines_model = moanin::create_splines_model(meta, degrees_of_freedom=6)
```

@@

The next step is to perform clustering. First, we reduce the set of genes of
interest to genes that are (1) significant; (2) "highly" differentially
expressed. First, we select the genes that have a log-fold change of at least
2 across at least one experiment for one time point. We thus use the object
`lfc_max`.

XXX

@@

kmeans_clusters = moanin:::splines_kmeans(
    y, splines_model, n_clusters=20,
    random_seed=42,
    n_init=20)

# Rescale the centroids
centroids = moanin:::rescale_values(kmeans_clusters$centroids, splines_model)

@@

```{r fig.height=8, fig.width=8}
moanin:::plot_centroids(centroids, splines_model,
			colors=as.vector(unlist(ann_colors["Group"])))
```
